What We Eval in the Shadows

### 2021-04-12 paper/main.tex
minor

The expressions passed to \eval vary widely. In order to categorize them, let us
use a minimization function $min(e)$ which for a given expression $e$ returns a
normal form that abstracts incidental details allowing the reader to focus on
the structure of the evaluated code. The minimization function performs constant
folding of arithmetic and string expressions for base operators, e.g.
$min(\c{1+1})=\c{V}$, value simplification, $min(\c{c(1,2,3+2)})=\c{V}$,
variable absorption, $min(\c{x+y})=\c{X}$, function absorption,
$min(\c{g(f(x),h(z))})=\c{F(F(X))}$ and a number of other simplifications.
Table~\ref{tab:minimizedexpressions} gives the 10 most frequent forms; \#sites
and \%sites are, respectively, the number and ratio of sites receiving arguments
of that form; \#packages is the number of packages with that form; \#operations
is the median number of instructions performed by the interpreter; and \%envir
is the percentage of sites for which the \c{envir} argument is current frame or one of the parent frames. The example column shows one sample
expression $e$ that is normalized to the particular form.

\begin{table}[h]\small
\begin{tabular}{|c|r|r|r|r|r|c|}\hline
  $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\\hline
\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\\hline
\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\\hline
\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\\hline
\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\\hline
\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\\hline
\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\\hline
\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\\hline
\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\\hline
\c{\{\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
\end{tabular}
\caption{Minimized expressions} \label{tab:minimizedexpressions}
\end{table}
% Other macros that can be used:
% - \packageMinimizedMedianOperations: median number of operations
% - \packageMinimizedPercentEnvir: percentages of sites with the default argument
% - \packageMinimizedPercentCallsEnvir: percentage of calls with the default envir argument

@@

\medskip\noindent\EE{$min(e)=\c{V}$} Expressions that represent values are
frequently passed to \eval -- they occur in 17\% of sites. The majority of
those, \packageValOneNodePercent, are inline constants (integer or double
vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
long as base functions such as \c{+} are not redefined. They typically aren't,
but this is a limitation of this categorization.} In our corpus,
\packageNbCallSitesUniqueActualValue call sites only ever see a simple value.
Manual inspection reveals cases such as the following
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
where the argument to \eval is a string which \eval simply returns. From twenty
randomly selected \evals, \packageUsefulValueEvalPercent of them do not need
\eval. For the other cases, either we are dealing with a value that needs to be
constructed dynamically or the value is a default case that sometimes is
replaced by a more interesting expression. This form is usually evaluated in few
interpreter steps, in fact the median is only
\packageMinimizedmedianoperationscRnd.

@@

>>>>

The expressions passed to \eval vary widely. In order to categorize them, let us
use a minimization function $min(e)$ which for a given expression $e$ returns a
normal form that abstracts incidental details allowing the reader to focus on
the structure of the evaluated code. The minimization function performs constant
folding of arithmetic and string expressions for base operators, e.g.
$min(\c{1+1})=\c{V}$, value simplification, $min(\c{c(1,2,3+2)})=\c{V}$,
variable absorption, $min(\c{x+y})=\c{X}$, function absorption,
$min(\c{g(f(x),h(z))})=\c{F(F(X))}$ and a number of other simplifications.
Table~\ref{tab:minimizedexpressions} gives the 10 most frequent forms; \#sites
and \%sites are, respectively, the number and ratio of sites receiving arguments
of that form; \#packages is the number of packages with that form; \#operations
is the median number of instructions performed by the interpreter; and \%envir
is the ratio of sites for that evaluate in a function environment. The example
column shows one sample expression $e$ that normalizes to the particular form.

\begin{table}[h]\small
\begin{tabular}{|c|r|r|r|r|r|c|}\hline
  $min(e)$& \#sites & \%sites & \#packages & \#operations & \%envir & example\\\hline
\c X&\packageMinimizedcallsitesa &\packageMinimizedpropsitesa &\packageMinimizedpackagea &\packageMinimizedmedianoperationsaRnd &\packageMinimizedpercentparentframesa & \c{y+1}\\\hline
\c{F(F(X))} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedmedianoperationsbRnd & \packageMinimizedpercentparentframesb & \c{gbov( mean(x), a-1)}\\\hline
\c{V}&\packageMinimizedcallsitesc &\packageMinimizedpropsitesc &\packageMinimizedpackagec &\packageMinimizedmedianoperationscRnd &\packageMinimizedpercentparentframesc& \c{c(42,21,0)}\\\hline
\c{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedmedianoperationsdRnd & \packageMinimizedpercentparentframesd & \c{seq\_len(iters)} \\\hline
\c{\$} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedmedianoperationseRnd & \packageMinimizedpercentparentframese & \c{DF\$B}\\\hline
\c{model.frame}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedmedianoperationsfRnd & \packageMinimizedpercentparentframesf &  \c{model.frame(formula = Z $\sim$ U)}   \\\hline
\c{F()}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedmedianoperationsgRnd & \packageMinimizedpercentparentframesg & \c{rgamma(3, 2, n = 10L)} \\\hline
\c{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedmedianoperationshRnd & \packageMinimizedpercentparentframesh & \c{function(x, y) x + 3 * y} \\\hline
\c{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedmedianoperationsiRnd & \packageMinimizedpercentparentframesi & \c{x[1, 2:3, 2:3] <- value}\\\hline
\c{BLOCK} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedmedianoperationsjRnd & \packageMinimizedpercentparentframesj & \c{\{ write.csv(iris, tf) ; file.size(tf) \}} \\\hline
\end{tabular}
\caption{Minimized expressions} \label{tab:minimizedexpressions}
\end{table}
% Other macros that can be used:
% - \packageMinimizedMedianOperations: median number of operations
% - \packageMinimizedPercentEnvir: percentages of sites with the default argument
% - \packageMinimizedPercentCallsEnvir: percentage of calls with the default envir argument

@@

\medskip\noindent\EE{$min(e)=\c{V}$} Expressions that represent values are
frequently passed to \eval -- they occur in 17\% of sites. The majority of
those, \packageValOneNodePercent, are inline constants (integer or double
vectors). The rest trivially evaluate to a value, \eg~\c{1+1}.\footnote{True as
long as base functions such as \c{+} are not redefined. They typically aren't,
but this is a limitation of this categorization.} In our corpus,
\packageNbCallSitesUniqueActualValue call sites only ever see a simple value.
Manual inspection reveals cases such as the following
\begin{lstlisting}
 f1 <- eval(paste("A~",paste(paste(names(X[,-1])),collapse="+")))
\end{lstlisting}
where the argument to \eval is a string which \eval simply returns. From twenty
randomly selected \evals, \packageUsefulValueEvalPercent of them do not need
\eval. For the other cases, either we are dealing with a value that needs to be
constructed dynamically or the value is a default case that sometimes is
replaced by a more interesting expression. This form is usually evaluated in few
interpreter steps, in fact the median is only
\packageMinimizedmedianoperationscRnd. The environment in which they evaluate is
mostly irrelevant (unless it redefines a built-in operator).
%% Why then this variety of environments??

@@

A more high-level classification: Table~\ref{tab:highlevelenvironments}

\begin{table}
    \centering
\begin{tabular}{|c|c|c|}
    \hline
    Environment & \#sites & \%sites \\
    \hline
    Parent frames &  \packageNbCallerEnvSites &  \packageCallerEnvSitePercent\\
    \hline
    Global &  \packageNbGlobalEnvSites & \packageNbGlobalEnvSitePercent \\
    \hline
    New & \packageNbNewEnvSites & \packageNewEnvSitePercent \\
    \hline
    Base &  \packageNbBaseEnvSites & \packageNbBaseEnvSitePercent \\
    \hline
    % Should be many more (next run)
    Package/namespace & \packageNbPackageEnvSites & \packagePackageEnvSitePercent \\
    \hline
    List &  \packageNbEnvListSites & \packageNbEnvListSitePercent \\
    \hline
    Integer & \packageNbIntegerEnvirSites  & \packageNbIntegerEnvirSitePercent \\
    \hline
    Null &  \packageNbNullEnvirSites & \packageNbNullEnvirSitePercent \\
    \hline
    Empty & \packageNbEmptyEnvSites & \packageNbEmptyEnvSitePercent \\
    \hline
\end{tabular} \label{tab:highlevelenvironments}
\caption{Classification of environments per site.}
\end{table}
